Experience Report: Haskell in Computational Biology

### 2012-07-02 mrfy_experience_report.tex
major restructuring of section 3.1 (models and viterbi)

Once a protein alignment is constructed, it~is used to train a
\emph{hidden Markov model}. 
A~hidden Markov model is a probabilistic finite-state machine which 
can assign a probability to any query sequence.
A~protein whose query sequence has a higher probability is more likely to %
be homologous to the proteins in the alignment.
We~write a query sequence as $x_1, \ldots, x_{\scriptscriptstyle N}$,
where each $x_i$~is 
an amino acid.
$N$~need not be the same as the number of columns in the alignment,
which we write~\alignwidth.

@@

A hidden Markov model also has distinguished \emph{Begin} and \emph{End} states.
In~our representation, each state contains a probability or a table of
probabilities, and it is also labeled with one of these labels:
\smallverbatiminput{statelabel}

@@

The computation is specified on the left-hand side of \figref{viterbi}.
A~probability $V_j^M(i)$ represents the probability of the most
likely path of the first~$i$ amino acids in the query sequence,
terminating with placement of amino acid $x_i$ in state~$M_j$.
Probabilities $V_j^D(i)$ and $V_j^D(i)$ are similar.
These
equations are very clearly explained by
\citet[Chapter~5]{Durbin:1998wz}

@@

Our code represents a query sequence as an immutable vector of amino
acids.
In~idiomatic Haskell, 
we might represent an individual amino acid~$x_i$
using a value of algebraic data type:
\begin{smallverbatim}
data Amino = Ala | Cys | Asp | Glu | Gly | ...
\end{smallverbatim}
But our models use a
legacy file format that represents each amino acid as a small integer.
Our representation is therefore
\smallverbatiminput{aa}

\remark{I'm not sure if this is the whole story, since the reader might simply
  ask, ``Why not map 0 to Ala, 1 to Cys, etc.?''. The real reason we keep the
  ``Int'' here I think, is that it provides easy indexing into the match and
  insertion emission vectors.}

The log of a probability is a negative number.
To~make these numbers more pleasant to read, we negate them,
which eliminates hordes of minus signs from our file formats.
The negated logarithm of a probability is called a \emph{score}.
\smallverbatiminput{score}
Type \texttt{Score} has a limited \texttt{Num} instance which permits
scores to be added and subtracted but not multiplied.

In a hidden Markov model, each transition is actually labeled with a
score, not a probability.
And in~a Plan7 model, as discussed above, 
there are 7~transitions into each node,
each labeled $a_{s\hat s}$,
where $s$~and~$\hat s$ form one of the 7~permissible pairs of state
labels: 
\smallverbatiminput{tprob-tprobs}
%\smallverbatiminput{tprobs}
To~choose a transition score from this 7-tuple, we use
function \texttt{aScore}, 
where
\mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {\hat s}$}.
A~7-tuple is stored in each node in
the model.
Each node also contains tables of emission probabilities
$e_{M_j}$~and~$e_{I_j}$, which are read using function
\mbox{$\mathtt{eScore}\;s\;j\;i = e_{s_j}(x_i)$}.

@@

In~Viterbi's equations,
the probability in each state is a function of the probabilities
in its predecessor states, 
and all probabilities can be computed by a classic dynamic-programming
algorithm.
This algorithm starts at the begin state,
computes probabilites in nodes $1$~through~$\alignwidth$ in
succession, and terminates at the end state.
One of us implemented this algorithm, storing the probabilities in an array.
The cost was
$O(|N|\times|\alignwidth|)$;
in~MRFy, $\alignwidth$~and~$N$ range from several hundred to a few
thousand.

@@

Another of us was curious to try coding Viterbi's equations
directly as recursive functions.
Like a recursive Fibonacci function, Viterbi's functions,
when implemented \naive ly,
take exponential time.
But like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
Our code implements a transformed version of Viterbi's equations, 
which operates on scores, not on probabilities.
This version is
shown on the right-hand side of \figref{viterbi}.
The transformed equations minimize the
score (the negated log probability) for each combination of
node~$j$, amino 
acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.

Scores can be usefully paired with many types of values,
so we have defined a small abstraction:
\smallfuzzverbatiminput{10.8pt}{vscore}
Think of a value of type \mono{Scored~a} as a container holding
 an~``\texttt a'' with a score written on the side.
The \texttt{/+/} function adds to the score without touching the
 container.

We~also made \texttt{Scored} an instance of~\texttt{Ord}.
Containers are ordered by score alone, so applying
\texttt{minimum} to a list of scored things chooses the thing with the
smallest (and therefore best) score.

@@

The \texttt{Scored} abstraction makes it easy to implement
Viterbi's equations.
For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
right of \figref{viterbi}, we call 
\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}.
The equation adds~$e^{\prime}_{M_{j}}(x_{i})$, computed
with \texttt{eScore}, to a minimum of sums.
The sum of an~$a'_{s\hat s}$~term and a $V^{\prime s}_{j-1}(i-1)$ term 
is computed by function \texttt{avSum}, in which
the terms are computed by \texttt{aScore} and \texttt{vee''}, respectively:
\smallverbatiminput{vfix}
%%%%%%%%%% horrible line break
% The right-hand side of \verb+vee'+ is wrapped in a call to
What about the call to
\mono{\mbox{fmap (Mat `cons`)}}?
This call performs a computation \emph{not} shown in \figref{viterbi}:
\mrfy\ computes not only the {probability} of the most likely path but 
also the actual path itself.
Function \mono{(Mat `cons`)} adds~$M$ to a path;
we~avoid \mbox{\mono{(Mat :)}} for reasons explained
in \secref{performance} below.

Function~\verb+vee''+ is the \emph{memoized} version of~\verb+vee'+.
Calling~\verb+vee''+ produces the same result as calling~\verb+vee'+,
but faster: 
\smallverbatiminput{memo}
Functions \texttt{Memo.memo3} and \texttt{Memo.arrayRange} come from
Luke Palmer's
\path{Data.MemoCombinators} package.
The value
\texttt{numNodes} represents~$\alignwidth$,
and \texttt{seqlen} represents~$N$.

>>>>

Once a protein alignment is constructed, it~is used to train a
\emph{hidden Markov model}. 
A~hidden Markov model is a probabilistic finite-state machine which 
can assign a probability to any query sequence.
A~protein whose query sequence has a higher probability is more likely to %
be homologous to the proteins in the alignment.
We~write a query sequence as $x_1, \ldots, x_{\scriptscriptstyle N}$,
where each $x_i$~is 
an amino acid.
The number of amino acids $N$ can differ from the number of columns in the alignment,
which we write~\alignwidth.

@@

A hidden Markov model also has distinguished ``begin'' and ``end''
states.
%%
%% for consistency with the text above 'begin' and 'end' are neither
%% capitalized nor italicized.
%%
In~our representation, each state contains a probability or a table of
probabilities, and it is also labeled with one of these labels:
\smallverbatiminput{statelabel}

@@

The computation is specified on the left-hand side of \figref{viterbi}.
A~probability $V_j^M(i)$ represents the probability of the most
likely path of the first~$i$ amino acids in the query sequence,
terminating with placement of amino acid $x_i$ in state~$M_j$.
Probabilities $V_j^D(i)$ and $V_j^D(i)$ are similar.
These
equations are explained very clearly by
\citet[Chapter~5]{Durbin:1998wz}

@@

Our code represents a query sequence as an immutable vector of amino
acids.
In~idiomatic Haskell, 
we might represent an individual amino acid~$x_i$
using a value of algebraic data type:
\begin{smallverbatim}
data Amino = Ala | Cys | Asp | Glu | ...   -- not used
\end{smallverbatim}
But our models use a
legacy file format in which each amino acid is a small integer
used only to index arrays.
%which we use only as an array index.
We therefore chose 
\smallverbatiminput{aa}

%%  \remark{I'm not sure if this is the whole story, since the reader might simply
%%    ask, ``Why not map 0 to Ala, 1 to Cys, etc.?''. The real reason we keep the
%%    ``Int'' here I think, is that it provides easy indexing into the match and
%%    insertion emission vectors.}

Our legacy file format also represents
the log of a probability \emph{negated}, which makes it a positive number.
%%To~make these numbers more pleasant to read, we negate them,
%%which eliminates hordes of minus signs from our file formats.
The negated logarithm of a probability is called a \emph{score}.
\smallverbatiminput{score}
Type \texttt{Score} has a limited \texttt{Num} instance which permits
scores to be added and subtracted but not multiplied.

In a real hidden Markov model, each probability is represented as a score.
Our code implements a transformed version of Viterbi's equations, 
which operates on scores, not on probabilities.
This version is
shown on the right-hand side of \figref{viterbi}.
The transformed equations minimize the
score (the negated log probability) for each combination of
column~$j$, amino 
acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.

A~model is represented as a sequence of \emph{nodes};
node~$j$ includes states $M_j$, $I_j$, and~$D_j$,
as well as the probabilities of transitions into that node.
Each node contains tables of emission scores
$e'_{M_j}$~and~$e'_{I_j}$.
These table are read by function
\texttt{eScore}, which satisfies
the law % parallel structure for line breaking below
\mbox{$\mathtt{eScore}\;s\;j\;i = e'_{s_j}(x_i)$}.
%
We place the transition probabilities 
into a record in which each field is labeled
\texttt{$s$\_$\mskip 1mu\hat s$},
where $s$~and~$\hat s$ form one of the 7~permissible pairs of state
labels: 
\smallverbatiminput{tprob-tprobs}
These scores are read by
function \texttt{aScore}, 
which satisfies
the law % line breaking
\mbox{$\mathtt{aScore}\;s\;\hat s\;(j-1) = \txprobj s {\hat s}$}.

@@

Scores can be usefully attached to many types of values,
so we have defined a small abstraction:
\smallfuzzverbatiminput{10.8pt}{vscore}
Think of a value of type \mono{Scored~a} as a container holding
 an~``\texttt a'' with a score written on the side.
The \texttt{/+/} function adds to the score without touching the
 container.
%
We~also made \texttt{Scored} an instance of~\texttt{Ord}.
Containers are ordered by score alone, so applying
\texttt{minimum} to a list of scored things chooses the thing with the
smallest (and therefore best) score.

Armed with our models and with the \texttt{Scored} abstraction, we
attacked Viterbi's equations. 
The probability in each state is a function of the probabilities
in its predecessor states, 
and all probabilities can be computed by a classic dynamic-programming
algorithm.
This algorithm starts at the {begin} state,
computes probabilites in nodes $1$~through~$\alignwidth$ in
succession, and terminates at the {end} state.
One of us implemented this algorithm, storing the probabilities in an array.
The cost was
$O(|N|\times|\alignwidth|)$;
in~MRFy, $\alignwidth$~and~$N$ range from several hundred to a few
thousand.

@@

@@

Another of us was curious to try coding Viterbi's equations
directly as recursive functions.
Like a recursive Fibonacci function, Viterbi's functions,
when implemented \naive ly,
take exponential time.
But like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
right of \figref{viterbi}, we define
\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}.
The equation adds~$e^{\prime}_{M_{j}}(x_{i})$, computed
with \texttt{eScore}, to a minimum of sums.
The~sum of an~$a'_{s\hat s}$~term and a $V^{\prime s}_{j-1}(i-1)$ term 
is computed by function \texttt{avSum}, in which
the terms are computed by \texttt{aScore} and \texttt{vee''}, respectively:
\smallfuzzverbatiminput{2.9pt}{vfix}
%%%%%%%%%% horrible line break
% The right-hand side of \verb+vee'+ is wrapped in a call to
What about the call to
\mono{\mbox{fmap (Mat `cons`)}}?
This call performs a computation \emph{not} shown in \figref{viterbi}:
\mrfy\ computes not only the {probability} of the most likely path but 
also the path itself.
Function \mono{(Mat `cons`)} adds~$M$ to a path;
we~avoid \mbox{\mono{(Mat :)}} for reasons explained
in \secref{performance} below.

Function~\verb+vee''+ is the {memoized} version of~\verb+vee'+.
Calling~\verb+vee''+ produces the same result as calling~\verb+vee'+,
but faster: 
\smallverbatiminput{memo}
Functions \texttt{Memo.memo3} and \texttt{Memo.arrayRange} come from
Luke Palmer's
\path{Data.MemoCombinators} package.
The value
\texttt{numNodes} represents~$\alignwidth$,
and \texttt{seqlen} represents~$N$.
